---
title: "MBDS522 Data Project"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Income Mobility and Measures of Wellbeing

### Introduction

Over the past 50 years in the the United States there has been a downward trend in intergenrational income mobility.[^1] In 1940 there was 95 percent likelihood that a child would make more than their parent. Yet, just 40 years later, a child born in 1980 only had about a 50 percent likelihood of outearning their parent (Chetty, Raj, et al., 2017).Clearly, the likelihood of a child making more than their parent has sharply declined in the past half century. See graph below.[^2] Yet, the question remains, are children aware of the decreased likelihood of outearning their parents? The importance of this questions is highlighted by Seligman's theory of learned helplessness (1972). This theory predicts that people can learn that they are helpless to change their situation and as a result exhibit symptoms of depression. Applied to this situation, the learned helplessness theory predicts that if children learn that they are helpless in their quest to outearn their parents, they will be more likely to develop depressive symptoms.   

[^1]:For this paper intergenerational income mobility is defined as the likelihood that a child will outearn their parent by parent income percentile. This  metric uses the child's earnings percentile at age 30 compared to their parent's earnings percentile. So, for example, to compute the mobility score of the 1980 cohort, one would use their 2010 earnings percentile and compare these to their parent's earnings percentile.   
[^2]:Data used for this graph was taken from Opportunity Insights https://opportunityinsights.org/data/


In light of the above conversation, and with such a sharp decline in income mobility in the past half century, this paper seeks to better understand the relationship between income mobility and negative wellbeing indicators such as sucide rates, and and subjective happiness. Interestingly, Nikolaev and Burns (2014) find a connection between intergenerational mobility and subjective wellbeing. For their measures of mobility they analyze educational, social and income outcomes. They find the correlation between mobility and subjective wellbeing to be consistent throughout all measurements, but especially in regard to income mobility. If subjective wellbeing is a predictor of sucide, Moum (1996) finds a negative correlation between subjective wellbeing and sucide, then one would expect to find states with lower levels of income mobility to be correlated to higher rates of suicide.   

Opposite what might be expected, Clark et. Al (2009) find that co-workers that are paid less than their peers are more likely to have job satisfaction. They hypothesize that they project their future pay to be higher and are therefore happier. Projecting this onto society as a whole, those who see trends of immobility may be less satisfied as they do not see potential future earnings. Additionally, Chetty et al. (2014) predict that intergenerational mobility is about the same for those born in the 1970's as those children entering the labor market in 2013 in the US. Yet, because we do not have data for future earnings of children born in 2013 this is difficult to konw. In extension of this finding, Chetty et al. (2014) suggest that increases in income inequality further highlight the importance of the birth lottery. It should be mentioned that income inequality combined with decreased likelihood of mobility fit nicely into the learned helplessness theory.  

In summary, this paper's purpose is to explore the connection between intergenerational income mobility and wellness outcomes such as sucide rates, and subjective happiness. Due to the nature of the data, this analysis does not attempt to prove causation. Instead, this analysis strives to find correlational evidence for a connection between intergenerational mobility and wellbeing. My hypothesis is that there is a positive correlation between income mobility and happiness and a negative correlation between income mobility and suicide rate. As will be discussed later, these results are particularly informative for governmental and charitable organisations.

## Data Collection

The data I collected came from a variety of sources. The data regarding suicide rates was taken from the the Center for Disease Control and Prevention (CDC). The CDC reported suicide rates per 100,000 at the state level. The data in this project used for income mobility was taken from an organization called Opportunity Insights. Opportunity Insights is a platformed based in Harvard that collects and organizes data to be used to help people make informed decisions and carry out high level research. Happiness measures were taken from the Wallet Hub. The Wallet Hub combined dozens of incators in psychological, physical, and community happiness to create overall state happiness measurements. An explanation of the data used and their methology is available on their website. The data for median household income by state was collected from the Census Bureau and used a 3 year average. To control for racial differences between states I collected data about racial percentages at the state level. The easiest format of this information that I found was on Wikipedia but the tables information was pulled from the the U.S. Census Bureau. The easiest was I found to control for race was to create a non-white control that was made up of the combination of black and hispanic racial percentages.   


For this paper intergenerational income mobility is defined as the likelihood that a child will outearn their parent by parent income percentile. This  metric uses the child's earnings percentile at age 30 compared to their parent's earnings percentile. So, for example, to compute the mobility score of the 1980 cohort, one would use their 2010 earnings percentile and compare these to their parent's earnings percentile. 


## Methodology 

The research in this study uses both cross-sectional and longitudinal data to explore the relationship between declining rates of income mobility and measures of wellbeing. Longitudinal data is used to get a high level overview of trends in both income mobility and wellbeing in the United States. However, because there are so many confounding variables which result in changes in wellbeing, cross-sectional data is used to further investigate the relationship. With a greater ability to control for outside factors such as racial makeup and average income, cross-sectional analysis will compose the majority of the analysis. The cross-sectional data analysis will focus mainly on the United States but will use country to country comparisons for further evidence.   


## Results Longitudinal 

### Income Mobility Overview  

Income mobility in the United States has been trending downwards since 1940. A simple regression shows that on average for every additional year, income mobility drops by 0.9 percentage points with a p-value below the .001 level.   


```{r}
options(Encoding="UTF-8")
cohort<-read.csv("cohortmeans.csv")

scatter.smooth(cohort$cohort_mean~cohort$year,lpars = list(lwd = 3, lty = 3), main= "US Income Mobilty by Year", xlab = "Cohort Year", ylab = "Mean Income Mobility")

```

<center> <h4>Regressing Income Mobility by Year </h4> </center>

```{r}
options(Encoding="UTF-8")
cohort<-read.csv("cohortmeans.csv")
library(huxtable)
reg<-lm(cohort$cohort_mean~cohort$year)
huxreg(reg, coefs = c('cohortyear'= 'cohort$year', "Intercept"='(Intercept)' ))

```

### Suicide Overview  

In the United States suicide does not have a linear trend. A non-linear trend is expected as many factors contribute to suicide. These include things such as genetics, environment, weather, and cultural influences. Suicide in the United States decreased from the 1960's to the year 2000. Since 2000 there has been a sharp increase in suicides.

```{r}
options(Encoding="UTF-8")
suicide<-read.csv("SuicideUS.csv")

scatter.smooth(suicide$Suicide~suicide$Year, main="US Suicide per 100,000", xlab = "Year", ylab = "Suicide")

```

At a quick glance this appears to be evidence that there is not a strong connection between suicide and income mobility. The problem with this assumption is that suicide is measured by year and not by cohort. This means that a suicide in a given year should be attributed to the individual's cohort instead of to the year. Due to the difficulty of getting cohort level suicide information, this study can only use a rough estimates. While definitely not exact science, if one assigns every suicide to the cohort born 40 years earlier (middle aged individuals have the highest suicide rates), then the relationship between cohort year and suicide still declines to begin and has an incline at the end. A linear regression model is not attempted to estimate correlation because of it's non-linearity.   


```{r}
options(Encoding="UTF-8")
suicide<-read.csv("cohortsuicide.csv")

scatter.smooth(suicide$Suicide~suicide$Cohort, main="US Suicide per 100,000", xlab = "Cohort Year", ylab = "Suicide")

```

Interestingly, when directly comparing cohort intergenerational income mobility levels in the United States to suicide levels by cohort, a strong correlation appears. As previously stated, suicide by cohort is only an estimate; however, though unclear at low income mobility levels, a strong positive correlation is found at higher levels of income mobility. Counter to intuition, this would suggest that higher levels of income mobility are correlated with higher suicide. The regression model suggests that a one hundred percent increase in Income Mobility is correlated with 3.87 more suicides per 100,000.      

```{r}
options(Encoding="UTF-8")
cohort<-read.csv("cohortmeans.csv")
library(huxtable)

scatter.smooth(cohort$Suicide~cohort$cohort_mean, main= "Suicide by Income Mobility", xlab = "Income Mobility", ylab = "Suicide per 100,000")

```

<center> <h4>Regressing Suicide on Income Mobility </h4> </center>

```{r}
options(Encoding="UTF-8")
cohort<-read.csv("cohortmeans.csv")
library(huxtable)

reg<-lm(cohort$Suicide~cohort$cohort_mean)

huxreg(reg, coefs = c('Cohort Mobility'='cohort$cohort_mean','Intercept'='(Intercept)'))

```


### Happiness Overview

Happiness or life satisfaction has only started to be collected in the past few decades. As a result, it is difficult to compare longitudinal happiness trends in the United States. Currently, we only have data for 2005 to 2017. However, from the data we do have, happiness has been steadily declining in the United States. See graph below.

```{r}
options(Encoding="UTF-8")
happy<-read.csv("happiness.csv")

scatter.smooth(happy$Happiness~happy$Year, main= "Happiness by Year", xlab = "Year", ylab = "Happiness")

```

Similar to suicide rates, it is difficult to compare a given year's reported life satisfaction to a specific cohort because life satisfaction measurements are composed of a variety of ages. Yet, if life satisfaction is compared to income mobility by assuming that happiness of a given year belong to the cohort of 40 years before, a positive correlation is found. This is consistent with what was expected. However, the correlation is not significant and will be further explored through cross-sectional data.   

```{r}
options(Encoding="UTF-8")
library(huxtable)
cohort<-read.csv("cohortmeans.csv")

scatter.smooth(cohort$Happiness~cohort$cohort_mean, main= "Happiness by Income Mobility", xlab = "Income Mobility", ylab = "Happiness")
```

<center> <h4>Regressing Happiness on Income Mobility </h4> </center>

```{r}
options(Encoding="UTF-8")
library(huxtable)
cohort<-read.csv("cohortmeans.csv")

reg<-lm(cohort$Happiness~cohort$cohort_mean)

huxreg(reg, coefs = c('Income Mobility'='cohort$cohort_mean','Intercept'='(Intercept)'))

```

## Results Cross-Sectional 

To better understand the relationship between income mobility and wellbeing, this study uses cross-sectional data in the United States. This data exploits state by state variation in income mobility to see if there is correlation to state by state variation in wellbeing measurements. With more data and the ability to control for outside causal factors, this study relies more heavily on cross-sectional data. From a quick analysis of income mobility by state, it appears that income mobility is approximately normally distributed across states. 

```{r}
options(Encoding="UTF-8")
data<-read.csv("projdata.csv")

hist(data$Mobility, main= "Histogram of Income Mobility by State",xlab="Income Mobility",breaks=20)

```

### Suicide Cross-Sectional Results

Using multiple linear regressions with added controls, evidence suggests that income mobility is negatively correlated with suicide. This supports the learned helplessness theory as those who are exposed to a greater likelihood of outearning their parents are less likely to commite suicide. While not proving causation, this suggests that there is relationship between income mobility and suicide.  

#### First Regression

The first regression is a simple linear regression of suicide rates on income mobility. This is consistent with the hypothesis that income mobility is correlated with lower levels of suicide. Interpreting this result suggests that a 100 percent increase increase in income mobility on average will decrease the suicide rate per 100,000 by 12.5 individuals. 

#### Second Regression

The second regression attempts for differences in median income across states. The median income is reported in 10,000's. Interestingly, when controlling for median income, the effect of income mobility on suicide increases. Interpreting these results suggests that a 100 percent increase in income mobility, holding income constant, on average leads to a 16 individual decrease in the suicide rate per 100,000.

#### Third Regression

The third regression attempts to account for differences in race across states. This is of importance because white males are the most likely to attempt suicide. Therefore, differences in racial makeup could be driving suicide rates instead of differences in income mobility. This regression uses the percentage of african americans and hispanics to compute a nonwhite variable for every state. When controlling for race income mobility becomes even more negatively correlated with suicide rates. This result suggests that holding median income and race constant, increasing income mobility by 100 percent on average is correlated with a 29.735 decrease in suicide rate per 100,000 individuals.  

<center> <h4> Regressing Suicide on Income Mobility with Controls </h4> </center>
```{r}
options(Encoding="UTF-8")
library(huxtable)
data<-read.csv("projdata.csv")

fit<-lm(data$RATE~data$Mobility)
fit2<-lm(data$RATE~data$Mobility+data$Median.Income)
fit3<-lm(data$RATE~data$Mobility+data$Median.Income+data$AfrHis)

huxreg(fit, fit2,fit3, coefs = c('Mobility'='data$Mobility', 'Median Income'='data$Median.Income','NonWhite'='data$AfrHis','Intercept'='(Intercept)'))
```



```{r}
options(Encoding="UTF-8")
library(huxtable)
data<-read.csv("projdata.csv")
plot(data$RATE~data$Mobility, main="State Income Mobility and Suicide", xlab="Income Mobility",ylab="Suicide Rate per 100,000")
abline(lm(data$RATE~data$Mobility))
```


### Happiness Cross-Sectional Results

Happiness in this report is measured through a combination of emotional, work, and community happiness reported in samples done throughout the United States. Obviously individual happiness is very subjective and confounded by a variety of factors. However, before diving deeper into wholistic happiness, I plotted happiness reports by their individual categories. Interestingly, combined happiness measurements are positively correlated with happiness but negatively correlated at each specific measurement.   

```{r}
options(Encoding="UTF-8")
data<-read.csv("projdata.csv")
plot(data$Happiness~data$Mobility,  main="Happiness", xlab="Income Mobility", ylab = "Happiness Rating")

abline(lm(data$Happiness~data$Mobility))

plot(data$EmoPhy~data$Mobility, main="Emotional and Phsycial Happiness", xlab="Income Mobility", ylab = "Emotional and Physical Happiness")
abline(lm(data$EmoPhy~data$Mobility))

plot(data$Community~data$Mobility,  main="Community Happiness", xlab="Income Mobility", ylab = "Community Happiness")
abline(lm(data$Community~data$Mobility))

plot(data$Work~data$Mobility,  main="Employment Happiness Happiness", xlab="Income Mobility", ylab = "Employment Happiness")
abline(lm(data$Work~data$Mobility))
```


#### First Regression

The first regression is a regression of happiness on income mobility. Consistent with what was expected, income mobility is positively correlated with happiness levels in a given state. On average a 100 percent increase in income mobility is correlated with a 40 point increase in average reported happiness. 

#### Second & Third Regression

In the second and third regression the relationship between income mobility and happiness is explored while adding in controls for median income and race. Mobility becomes significant at the 5 percent level when adding in these controls. Yet, because of the many confounding factors such as culture, weather, etc., a direct causal relationship is not suspected. It is noteworthy that the sign stays consistent throughout these regressions pointing to income mobility being positively correlated to reported happiness. 

```{r}
options(Encoding="UTF-8")
data<-read.csv("projdata.csv")
library(huxtable)

fit<-lm(data$Happiness~data$Mobility)
fit2<-lm(data$Happiness~data$Mobility+data$Median.Income)
fit3<-lm(data$Happiness~data$Mobility+data$Median.Income+data$AfrHis)

huxreg(fit, fit2,fit3, coefs = c('Mobility'='data$Mobility', 'Median Income'='data$Median.Income','NonWhite'='data$AfrHis','Intercept'='(Intercept)'))



```


## Conclusion and Extensions

Admittedly, with hundreds of confounding factors it is impossible to conclude any sort of causation between income mobility and suicide rates or happiness levels. While it is impossible to conclude causation, there is intriguing evidence linking the two outcomes to changes in income mobility. Even if their is no causal link, I believe these findings to be of interest for further study and investigation. 

Interestingly, suicide rates held a negative correlation to income mobility through multiple controls.  Though not of strong significance, I believe that this supports the hypothesis that income mobility would be correlated with suicide rates. This was particularly convincing because year to year trends show suicide positively correlated to income mobility but variation in income mobility between states showed the expected negative correlation. The longitudinal suicide rates did not show a negative correlation to income mobility, but with so many confounding factors such as cultural influences, I believe the cross sectional data to be of more importance when analyzing this correlation.  

Happiness and income mobility showed promise of a positive correlation when looking at longitudinal data. Yet, these results became a little messier when looking at cross sectional data. First, income mobility had the opposite expected correlation with individual measurements of happiness. With each measurment income mobility was negatively correlated. This made the later finding that overall level of happiness to be positively correlated to happiness less strong. However, income mobility was consistently positively correlated to happiness accross multiple controls. Overall, I conclude that my evidence was not strong enough to support my hypothesis that income mobility is positively correlated to happiness within states. 

This information and the connection between income mobility and suicide/ happiness levels are important to goverment and charity organizations. By tackling the problems surrounding income mobility and helping people increase their likelihood of success there will likely be improvements in suicide and depression rates. I suggest further research with more data and more controls. 


***  

### Bibliography 
***
Chetty, Raj, Nathaniel Hendren, Patrick Kline, Emmanuel Saez, and Nicholas Turner. 2014. "Is the United States Still a Land of Opportunity? Recent Trends in Intergenerational Mobility." American Economic Review, 104 (5): 141-47.

Chetty, Raj, et al. "The fading American dream: Trends in absolute income mobility since 1940." Science 356.6336 (2017): 398-406.

Clark, Andrew E., Nicolai Kristensen, and Niels Westergård-Nielsen. "Job satisfaction and co-worker wages: Status or signal?." The Economic Journal 119.536 (2009): 430-447.

Moum, Torbjorn. "Subjective well-being as a short-and long-term predictor of suicide in the general population." World Conference in Quality of Life, University of Northern British Columbia, Prince George, Canada. August. 1996.

Nikolaev, Boris, and Ainslee Burns. "Intergenerational mobility and subjective well-being-Evidence from the general social survey." Journal of Behavioral and Experimental Economics53 (2014): 82-96.

Seligman, Martin EP. "Learned helplessness." Annual review of medicine 23.1 (1972): 407-412.

***
 

